\large \textbf{Burrows-Wheeler transform for terabases}

### 2015-10-29 paper/paper.tex
Merge example.

% FIXME: a running example with 2 texts
Assume that we have two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively, and that we have already built \BWT{}s for them. Assume that we store the \BWT{}s in a two-level array, where the first level contains pointers to \emph{blocks} of size $b$ (bits). If the \BWT{} takes $x$ bits, the two-level array imposes $\frac{x}{b} \log x + \Oh(b)$ bits of space overhead. With $b = \sqrt{x \log x}$, the overhead becomes $\Oh(\sqrt{x \log x})$ bits.

@@

\end{enumerate}
The overall time complexity of \BWT{} merging is $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$. The algorithm uses $n_{\Bcoll} \log n_{\Acoll} + \Oh(\sqrt{x \log x})$ bits of working space in addition to the \BWT{}s and the structures required to use them as FM-indexes.

@@

\smallbreak\noindent\textbf{Buffering.} To reduce disk usage and the amount of disk I/O, we buffer and compress the lexicographic ranks before writing them to disk. Each thread has two buffers: a \emph{run buffer} and a \emph{thread buffer}. The run buffer stores the runs as pairs of integers $(r, \ell)$. Once the run buffer becomes full, we sort the runs by \emph{run heads} $r$, use \emph{differential encoding} for the run heads, and encode the differences and run lengths with a \emph{prefix-free code}. The compressed run buffer is then merged with the similarly compressed thread buffer.

@@

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments. \BWT{} sizes are for the native format. RLO indicates that the reads are sorted in reverse lexicographic order in each input file.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|ccc|cc}
\hline
 & & & & & \multicolumn{2}{c}{\textbf{\BWT{} size}} \\
\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} & \textbf{Unsorted} & \textbf{RLO} \\
\hline
\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp & 50.3 GB \\
     & NA12891 &  242 Gbp & 2.40G &       101 bp & 42.4 GB \\
     & NA12892 &  245 Gbp & 2.42G &       101 bp & 43.8 GB \\
     & Merged  &  771 Gbp & 7.63G &       101 bp & \\
\hline
\RS  & AA      &  433 Gbp & 4.69G & 73 or 100 bp &      -- & 38.5 GB \\
     & TT      &  432 Gbp & 4.68G & 73 or 100 bp &      -- & 38.7 GB \\
     & AT      &  275 Gbp & 2.98G & 73 or 100 bp &      -- & 26.6 GB \\
     & TA      &  355 Gbp & 3.84G & 73 or 100 bp &      -- & 32.7 GB \\
     & Merged  & 1.49 Tbp & 16.2G & 73 or 100 bp &      -- &  117 GB \\
\hline
\RS  & *A, *C  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- &  225 GB \\
     & Merged  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- & \\
\hline
\RS  & *G, *T  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- &  226 GB \\
     & Merged  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- & \\
\hline
\end{tabular}}
\end{center}
\end{table}

>>>>

\begin{figure}[t!]
\includegraphics[width=\textwidth]{merge_example.pdf}
\caption{Merging the \BWT{}s of texts $R$ and $S$. Rank array \RA{} counts the number of suffixes of $R$ that are lexicographically smaller than or equal to the given suffix of $S$. We fill it by starting with $\mRA[1] = 1$ and iterating $\mRA[\mLF_{S}(i)] = \mLF_{R}(\mRA[i], \mBWT_{S}[i])$. Interleaving bitvector $B_{RS}$ tells whether the source of a character in the merged \BWT{} is in $\mBWT_{R}$ or $\mBWT_{S}$. We build it by setting bits $i+\mRA[i]$ to 1 for all $i$.}\label{fig:merge}
\end{figure}

Assume that we have two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively, and that we have already built \BWT{}s for them. Assume that we store the \BWT{}s in two-level arrays, where the first level contains pointers to \emph{blocks} of size $b$ (bits). If the \BWT{} takes $x$ bits, the two-level array imposes $\frac{x}{b} \log x + \Oh(b)$ bits of space overhead. With $b = \sqrt{x \log x}$, the overhead becomes $\Oh(\sqrt{x \log x})$ bits.

@@

\end{enumerate}
See Figure~\ref{fig:merge} for an example with two texts. The overall time complexity of \BWT{} merging is $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$. The algorithm uses $n_{\Bcoll} \log n_{\Acoll} + \Oh(\sqrt{x \log x})$ bits of working space in addition to the \BWT{}s and the structures required to use them as FM-indexes.

@@

\smallbreak\noindent\textbf{Buffering.} To reduce disk I/O and disk usage, we buffer and compress the lexicographic ranks before writing them to disk. Each thread has two buffers: a \emph{run buffer} and a \emph{thread buffer}. The run buffer stores the runs as pairs of integers $(r, \ell)$. Once the run buffer becomes full, we sort the runs by \emph{run heads} $r$, use \emph{differential encoding} for the run heads, and encode the differences and run lengths with a \emph{prefix-free code}. The compressed run buffer is then merged with the similarly compressed thread buffer.

@@

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments. \BWT{} sizes are for the native format. RLO indicates that the reads are sorted in reverse lexicographic order.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|ccc|cc}
\hline
 & & & & & \multicolumn{2}{c}{\textbf{\BWT{} size}} \\
\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} & \textbf{Unsorted} & \textbf{RLO} \\
\hline
\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp & 50.3 GB & 25.5 GB \\
     & NA12891 &  242 Gbp & 2.40G &       101 bp & 42.4 GB & 19.7 GB \\
     & NA12892 &  245 Gbp & 2.42G &       101 bp & 43.8 GB & 20.7 GB \\
     & Merged  &  771 Gbp & 7.63G &       101 bp &  129 GB & \\
\hline
\RS  & AA      &  433 Gbp & 4.69G & 73 or 100 bp &      -- & 38.5 GB \\
     & TT      &  432 Gbp & 4.68G & 73 or 100 bp &      -- & 38.7 GB \\
     & AT      &  275 Gbp & 2.98G & 73 or 100 bp &      -- & 26.6 GB \\
     & TA      &  355 Gbp & 3.84G & 73 or 100 bp &      -- & 32.7 GB \\
     & Merged  & 1.49 Tbp & 16.2G & 73 or 100 bp &      -- &  117 GB \\
\hline
\RS  & *A, *C  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- &  225 GB \\
     & Merged  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- & \\
\hline
\RS  & *G, *T  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- &  226 GB \\
     & Merged  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- & \\
\hline
\end{tabular}}
\end{center}
\end{table}
